Lesson 5
========================================================

### Multivariate Data
Notes:

***

### Moira Perceived Audience Size Colored by Age
Notes:

***

### Third Qualitative Variable
Notes:

```{r Third Qualitative Variable}
library(ggplot2)
pf <- read.csv('../lesson3/pseudo_facebook.tsv', sep = '\t')
ggplot(aes(x = gender, y = age),
       data = subset(pf, !is.na(gender))) + geom_boxplot() +
  stat_summary(fun.y = mean, geom = 'point', shape = 4)

ggplot(aes(x = age, y = friend_count),
       data = subset(pf, !is.na(gender))) + 
  geom_line(aes(color = gender), stat = 'summary', fun.y = median)
```

***

### Plotting Conditional Summaries
Notes:

```{r Plotting Conditional Summaries}
library(dplyr)
pf.fc_by_age_gender <- pf %>%
  filter(!is.na(gender)) %>% group_by(age, gender) %>% 
  summarize(mean_friend_count = mean(friend_count),
          median_friend_count = median(as.numeric(friend_count)),
          n = n()) %>%
  arrange(age, gender) %>% ungroup()
head(pf.fc_by_age_gender, 10)

ggplot(aes(x = age, y = median_friend_count), data = pf.fc_by_age_gender) +
  geom_line(aes(color = gender))
```

***

### Thinking in Ratios
Notes:

***

### Wide and Long Format
Notes:

***

### Reshaping Data
Notes:

```{r}
library(reshape2)
pf.fc_by_age_gender.wide <- dcast(pf.fc_by_age_gender,
                                  age ~ gender,
                                  value.var = 'median_friend_count')
head(pf.fc_by_age_gender.wide)

#We could also create a similar data frame using the dplyr and tidyr packages: 
#pf.fc_by_age_gender.wide <-     subset(pf.fc_by_age_gender[c('age', 'gender', 'median_friend_count')],                   !is.na(gender)) %>% 
#    spread(gender, median_friend_count) %>% 
#    mutate(ratio = male / female) 
#
#head(pf.fc_by_age_gender.wide)
```

***

### Ratio Plot
Notes:

```{r Ratio Plot}
ggplot(aes(x = age, y = female/ male), data = pf.fc_by_age_gender.wide) +
  geom_line() +
  geom_hline(aes(yintercept = 1), alpha = 0.3, linetype = 'dashed')
```

***

### Third Quantitative Variable
Notes:

```{r Third Quantitative Variable}
max(pf$tenure, na.rm = T)
pf$year_joined = floor(2014 - pf$tenure / 365)
table(pf$year_joined)
```

***

### Cut a Variable
Notes:

```{r Cut a Variable}
# You need to create the following buckets for the
# new variable, year_joined.bucket

#        (2004, 2009]
#        (2009, 2011]
#        (2011, 2012]
#        (2012, 2014]

pf$year_joined.bucket = cut(pf$year_joined, c(2004, 2009, 2011, 2012, 2014))

```

***

### Plotting it All Together
Notes:

```{r Plotting it All Together}
table(pf$year_joined.bucket)
table(pf$year_joined.bucket, useNA = 'ifany')

ggplot(aes(x = age, y = friend_count),
       data = subset(pf, !is.na(year_joined.bucket))) + 
  geom_line(aes(color = year_joined.bucket), stat = 'summary', fun.y = median)
```

***

### Plot the Grand Mean
Notes:

```{r Plot the Grand Mean}
ggplot(aes(x = age, y = friend_count),
       data = subset(pf, !is.na(year_joined.bucket))) + 
  geom_line(aes(color = year_joined.bucket), stat = 'summary', fun.y = mean) +
  geom_line(stat = 'summary', linetype = 'dashed', fun.y = mean)

```

***

### Friending Rate
Notes:

```{r Friending Rate}
rate = with(subset(pf, tenure > 0), friend_count / tenure)
median(rate)
max(rate)

```

***

### Friendships Initiated
Notes:

What is the median friend rate?

What is the maximum friend rate?

```{r Friendships Initiated}
ggplot(aes(x = tenure, y = friendships_initiated / tenure),
       data = subset(pf, tenure > 0 & !is.na(year_joined.bucket))) + 
  geom_line(aes(color = year_joined.bucket), stat = 'summary', fun.y = mean) #+
  #geom_line(stat = 'summary', linetype = 'dashed', fun.y = mean)

ggplot(aes(x = 7 * round(tenure / 7), y = friendships_initiated / tenure),
       data = subset(pf, tenure > 0 & !is.na(year_joined.bucket))) + 
  geom_line(aes(color = year_joined.bucket), stat = 'summary', fun.y = mean) #+


ggplot(aes(x = tenure, y = friendships_initiated / tenure),
       data = subset(pf, tenure > 0)) +
  geom_smooth(aes(color = year_joined.bucket)) 
```

***

### Bias-Variance Tradeoff Revisited
Notes:

```{r Bias-Variance Tradeoff Revisited}

ggplot(aes(x = tenure, y = friendships_initiated / tenure),
       data = subset(pf, tenure >= 1)) +
  geom_line(aes(color = year_joined.bucket),
            stat = 'summary',
            fun.y = mean)

ggplot(aes(x = 7 * round(tenure / 7), y = friendships_initiated / tenure),
       data = subset(pf, tenure > 0)) +
  geom_line(aes(color = year_joined.bucket),
            stat = "summary",
            fun.y = mean)

ggplot(aes(x = 30 * round(tenure / 30), y = friendships_initiated / tenure),
       data = subset(pf, tenure > 0)) +
  geom_line(aes(color = year_joined.bucket),
            stat = "summary",
            fun.y = mean)

ggplot(aes(x = 90 * round(tenure / 90), y = friendships_initiated / tenure),
       data = subset(pf, tenure > 0)) +
  geom_line(aes(color = year_joined.bucket),
            stat = "summary",
            fun.y = mean)

```

***

### Sean's NFL Fan Sentiment Study
Notes:

***

### Introducing the Yogurt Data Set
Notes:

***

### Histograms Revisited
Notes:

```{r Histograms Revisited}
yo <- read.csv("yogurt.csv")
str(yo)
yo$id <- factor(yo$id)
library(ggplot2)
ggplot(aes(x = price), data = yo) +
  geom_histogram()
```

***

### Number of Purchases
Notes:

```{r Number of Purchases}
unique(yo$price)
length(unique(yo$price))
table(yo$price)

yo = transform(yo, all.purchases = strawberry + blueberry + pina.colada + plain + mixed.berry)
ggplot(aes(x = all.purchases), data = yo) +
  geom_histogram(binwidth = 1)
```

***

### Prices over Time
Notes:

```{r Prices over Time}
ggplot(aes(x = time, y = price), data = yo) +
  geom_jitter(alpha = 1/20)

```

***

### Sampling Observations
Notes:

***

### Looking at Samples of Households

```{r Looking at Sample of Households}
set.seed(4230)
sample.ids = sample(levels(yo$id), 16)

ggplot(aes(x = time, y = price), 
       data = subset(yo, id %in% sample.ids)) +
  facet_wrap( ~ id) + 
  geom_line() +
  geom_point(aes(size = all.purchases), pch = 1)
```

***

### The Limits of Cross Sectional Data
Notes:

***

### Many Variables
Notes:

***

### Scatterplot Matrix
Notes:
```r{}
library(GGally)
theme_set(theme_minimal(20))
set.seed(1836)
pf_subset <- pf[, c(2:15)]
names(pf_subset)
#ggpairs(pf_subset[sample.int(nrow(pf_subset), 1000), ])

#You may also find that variable labels are on the outer edges of the scatterplot
#matrix, rather than on the diagonal. If you want labels in the diagonal, you can set
#the axisLabels = 'internal' argument in your ggpairs command.
```
***

### Even More Variables
Notes:

***

### Heat Maps
Notes:

```{r}
nci <- read.table("nci.tsv")
colnames(nci) <- c(1:64)
```

```{r}
library(reshape2)
nci.long.samp <- melt(as.matrix(nci[1:200,]))
names(nci.long.samp) <- c("gene", "case", "value")
head(nci.long.samp)

ggplot(aes(y = gene, x = case, fill = value),
  data = nci.long.samp) +
  geom_tile() +
  scale_fill_gradientn(colours = colorRampPalette(c("blue", "red"))(100))
```


***

### Analyzing Three of More Variables
Reflection:

***

Click **KnitHTML** to see all of your hard work and to have an html
page of this lesson, your answers, and your notes!

