Lesson 4
========================================================

***

### Scatterplots and Perceived Audience Size
Notes:

***

### Scatterplots
Notes:

```{r Scatterplots}
library(ggplot2)
pf <- read.csv('../lesson3/pseudo_facebook.tsv', sep = '\t')
#You can also read in the data using the following code: 
#read.delim('pseudo_facebook.tsv')
names(pf)
#names(pf)[c(-2, -1)]
qplot(x = age, y = friend_count, data = pf)
```

***

#### What are some things that you notice right away?
Response:

In general, it is a good idea to think about what caused certain oddities or values in data sets. This can help explain dirty data.
***

### ggplot Syntax
Notes:

```{r ggplot Syntax}
#The ggplot syntax for the scatterplot: 
ggplot(aes(x = age, y = friend_count), data = pf) + 
  geom_point() +
  xlim(13, 90)

```

***

### Overplotting
Notes:

```{r Overplotting}
ggplot(aes(x = age, y = friend_count), data = pf) + 
  geom_point(alpha = 1/20) +
  xlim(13, 90)

ggplot(aes(x = age, y = friend_count), data = pf) + 
  geom_jitter(alpha = 1/20) +
  xlim(13, 90)
```

#### What do you notice in the plot?
Response:

***

### Coord_trans()
Notes:
Look up the documentation for coord_trans() by running this line of code ?coord_trans or visit http://docs.ggplot2.org/current/coord_trans.html.

```{r Coord_trans()}
ggplot(aes(x = age, y = friend_count), data = pf) + 
  geom_point(alpha = 1/20, position = position_jitter(h = 0)) +
  coord_trans(y = "sqrt") +
  xlim(13, 90) 

#sqrt(pf$friend_count)
```

#### Friends initiated vs. age

```{r}
names(pf)
ggplot(aes(x = age, y = friendships_initiated), data = pf) +
  geom_point(alpha = 1/20, position = position_jitter(h = 0)) + 
  coord_trans(y = "sqrt")
```

#### What do you notice?

***

### Alpha and Jitter
Notes:

```{r Alpha and Jitter}

```

***

### Overplotting and Domain Knowledge
Notes:

***

### Conditional Means
Notes:
Another warning: Version 0.4 of R has a bug when using the median function on the summarize layer, depending on the nature of the data being summarized. You may need to cast the data as a numeric (float) type when using it on your local machine, e.g. median(as.numeric(var)).
There are other ways to work with data and create new data frames without using the dplyr package. Learn about the R functions lapply, tapply, and split in a [blog post]( http://rollingyours.wordpress.com/2014/10/20/the-lapply-command-101/) 

```{r Conditional Means}
library(dplyr)
age_groups = group_by(pf, age)
pf.fc_by_age = summarize(age_groups, 
          friend_count_mean = mean(friend_count),
          friend_count_median = median(friend_count),
          n = n())
pf.fc_by_age = arrange(pf.fc_by_age, age)
head(pf.fc_by_age)
```

```{r Conditional Means, Alternative}
library(dplyr)
pf.fc_by_age <- pf %>% group_by(age) %>% 
  summarize(friend_count_mean = mean(friend_count),
          friend_count_median = median(friend_count),
          n = n()) %>%
  arrange(age)
head(pf.fc_by_age, 10)
```

Create your plot!

```{r Conditional Means Plot}
# Plot mean friend count vs. age using a line graph.
ggplot(aes(x = age, y = friend_count_mean), data = pf.fc_by_age) +
  geom_line()
```

***

### Overlaying Summaries with Raw Data
Notes:

```{r Overlaying Summaries with Raw Data}
ggplot(aes(x = age, y = friend_count), data = pf) +
  geom_point(alpha = 1/20, position = position_jitter(h = 0), 
             color = 'orange') + 
  #coord_trans(y = "sqrt") +
  coord_cartesian(xlim = c(13, 90), ylim = c(0, 1000)) +
  geom_line(stat = 'summary', fun.y = mean) +
  geom_line(stat = 'summary', fun.y = quantile, probs = .1,
            linetype = 2, color = 'blue') +
  geom_line(stat = 'summary', fun.y = quantile, probs = .5,
            color = 'blue') +
  geom_line(stat = 'summary', fun.y = quantile, probs = .9,
            linetype = 2, color = 'blue') 
  

```

#### What are some of your observations of the plot?
Response:

***

### Moira: Histogram Summary and Scatterplot
See the Instructor Notes of this video to download Moira's paper on perceived audience size and to see the final plot.

Notes:

***

### Correlation
Notes:

```{r Correlation}
cor(pf$friend_count, pf$age)
cor.test(pf$friend_count, pf$age)
with(pf, cor.test(friend_count, age, method = 'pearson'))
```

Look up the documentation for the cor.test function.

What's the correlation between age and friend count? Round to three decimal places.
Response:

***

### Correlation on Subsets
Notes:

```{r Correlation on Subsets}
#with(                 , cor.test(age, friend_count))
```

***

### Correlation Methods
Notes:

***

## Create Scatterplots
Notes:

```{r}
ggplot(aes(x = www_likes_received, y = likes_received), data = pf) +
  coord_trans(y = "sqrt", x = "sqrt") + 
  geom_point()  

ggplot(aes(x = www_likes_received, y = likes_received), data = pf) +
  geom_point() +
  xlim(0, quantile(pf$www_likes_received, 0.95)) +
  ylim(0, quantile(pf$likes_received, 0.95)) +
  geom_smooth(method = 'lm', color = 'red')
```

***

### Strong Correlations
Notes:

```{r Strong Correlations}

```

What's the correlation betwen the two variables? Include the top 5% of values for the variable in the calculation and round to 3 decimal places.

```{r Correlation Calcuation}
with(pf, cor(www_likes_received, likes_received))
```

Response:

***

### Moira on Correlation
Notes:

***

### More Caution with Correlation
Notes:

```{r More Caution With Correlation}
#install.packages('alr3')
library(alr3)
data("Mitchell")
```

Create your plot!

```{r Temp vs Month}
ggplot(aes(x = Month, y = Temp), data = Mitchell) +
  geom_point()
```

***

### Noisy Scatterplots
a. Take a guess for the correlation coefficient for the scatterplot.
0.1
b. What is the actual correlation of the two variables?
(Round to the thousandths place)
```{r}
with(Mitchell, cor.test(Month, Temp))
```
0.057

```{r Noisy Scatterplots}
```

***

### Making Sense of Data
Notes:

```{r Making Sense of Data}
ggplot(aes(x = Month, y = Temp), data = Mitchell) +
  geom_point() +
  scale_x_discrete(breaks = seq(0, max(Mitchell$Month), 12))
 
ggplot(aes(x=(Month%%12),y=Temp),data=Mitchell)+ 
  geom_point() 

```

***

### A New Perspective

What do you notice?
Response:
oscillations

Watch the solution video and check out the Instructor Notes!
Notes:
There are other measures of associations that can detect this. The dcor.ttest() function in the 'energy' package implements a non-parametric test of the independence of two variables. The test correctly rejects the independence.
***

### Understanding Noise: Age to Age Months
Notes:

```{r Understanding Noise: Age to Age Months}
ggplot(aes(x = age, y = friend_count_mean), data = pf.fc_by_age) +
  geom_line()

```

***

### Age with Months Means

```{r Age with Months Means}
pf$age_with_months <- pf$age + (12 - pf$dob_month) / 12
```

Programming Assignment
```{r Programming Assignment}
pf.fc_by_age_months <- pf %>% group_by(age_with_months) %>% 
  summarize(friend_count_mean = mean(friend_count),
          friend_count_median = median(friend_count),
          n = n()) %>%
  arrange(age_with_months)
head(pf.fc_by_age_months, 10)

```

***

### Noise in Conditional Means

```{r Noise in Conditional Means}
p1 <- ggplot(aes(x = age, y = friend_count_mean), data = subset(pf.fc_by_age, age < 71)) +
  geom_line() +
  geom_smooth()

p2 <- ggplot(aes(x = age_with_months, y = friend_count_mean), data =
         subset(pf.fc_by_age_months, age_with_months < 71)) +
  geom_line() +
  geom_smooth()

p3 <- ggplot(aes(x = round(age / 5) * 5, y = friend_count_mean), 
             data = subset(pf.fc_by_age, age < 71)) +
  geom_line(stat = 'summary', fun.y = 'mean')
library(gridExtra)
grid.arrange(p1, p2, p3, ncol = 1)
```

***

### Smoothing Conditional Means
Notes:

```{r Smoothing Conditional Means}
ggplot(aes(x = age_with_months, y = friend_count_mean), data =
         subset(pf.fc_by_age_months, age_with_months <= 71)) +
  geom_line()

```

***

### Which Plot to Choose?
Notes:

***

### Analyzing Two Variables
Reflection:

***

Click **KnitHTML** to see all of your hard work and to have an html
page of this lesson, your answers, and your notes!

