Lesson 3
========================================================

***


### What to Do First?
Notes:
```{r}
getwd()
list.files()
```

***

### Pseudo-Facebook User Data
Notes:

```{r Pseudo-Facebook User Data}
pf <- read.csv('pseudo_facebook.tsv', sep = '\t')
names(pf)
```

***

### Histogram of Users' Birthdays
Notes:

```{r Histogram of Users\' Birthdays}
#install.packages('ggplot2')
library(ggplot2)
qplot(data = pf, x = dob_day) +
  scale_x_discrete(breaks = 1:31)
ggplot(aes(x = dob_day), data = pf) + 
  geom_histogram() + 
  scale_x_discrete(breaks = 1:31)
```

***

#### What are some things that you notice about this histogram?
Response:

***

### Moira's Investigation
Notes:

***

### Estimating Your Audience Size
Notes:

***

#### Think about a time when you posted a specific message or shared a photo on Facebook. What was it?
Response:

#### How many of your friends do you think saw that post?
Response:

#### Think about what percent of your friends on Facebook see any posts or comments that you make in a month. What percent do you think that is?
Response:

***

### Perceived Audience Size
Notes:

***
### Faceting
Notes:

```{r Faceting}
qplot(data = pf, x = dob_day) +
  scale_x_discrete(breaks = 1:31) +
  facet_wrap(~dob_month, ncol = 3)

# facet_grid(vertical~horizontal)
```

#### Let’s take another look at our plot. What stands out to you here?
Response:

***

### Be Skeptical - Outliers and Anomalies
Notes:

***

### Moira's Outlier
Notes:
#### Which case do you think applies to Moira’s outlier?
Response:

***

### Friend Count
Notes:

#### What code would you enter to create a histogram of friend counts?

```{r Friend Count}
qplot(data = pf, x = friend_count)  
  #scale_x_discrete(breaks = 1:5000) 
```

#### How is this plot similar to Moira's first plot?
Response:

***

### Limiting the Axes
Notes:

```{r Limiting the Axes}

qplot(data = pf, x = friend_count, xlim = c(0, 1000))  

qplot(data = pf, x = friend_count) +
  scale_x_continuous(limits = c(0, 1000))
```

### Exploring with Bin Width
Notes:

***

### Adjusting the Bin Width
Notes:

```{r Adjusting the Bin Width}
qplot(data = pf, x = friend_count, binwidth = 25) +
  scale_x_continuous(limits = c(0, 1000), breaks = seq(0, 1000, 50))
```

### Faceting Friend Count
```{r Faceting Friend Count}
# What code would you add to create a facet the histogram by gender?
# Add it to the code below.
qplot(x = friend_count, data = pf, binwidth = 10) +
  scale_x_continuous(limits = c(0, 1000),
                     breaks = seq(0, 1000, 50)) +
  facet_wrap(~gender)
```

***

### Omitting NA Values
Notes:

```{r Omitting NA Values}
ggplot(aes(x = friend_count), data = subset(pf, !is.na(gender))) + 
  geom_histogram(binwidth = 25) + 
  scale_x_continuous(limits = c(0, 1000), breaks = seq(0, 1000, 50)) + 
  facet_wrap(~gender)

# omitting NA in any field
ggplot(aes(x = friend_count), data = na.omit(pf)) + 
  geom_histogram(binwidth = 25) + 
  scale_x_continuous(limits = c(0, 1000), breaks = seq(0, 1000, 50)) + 
  facet_wrap(~gender)
```

***

### Statistics 'by' Gender
Notes:

```{r Statistics \'by\' Gender}
table(pf$gender)
by(pf$friend_count, pf$gender, summary)
```

#### Who on average has more friends: men or women?
Response:

#### What's the difference between the median friend count for women and men?
Response:

#### Why would the median be a better measure than the mean?
Response:
  Median is less sensitive to long tails and outliers
***

### Tenure
Notes:

```{r Tenure}
qplot(x = tenure, data = pf,
   color = I('black'), fill = I('#099DD9'))

ggplot(aes(x = tenure), data = pf) + 
   geom_histogram(binwidth = 30, color = 'black', fill = '#099DD9')
```

***

#### How would you create a histogram of tenure by year?

```{r Tenure Histogram by Year}
qplot(x = tenure/365, data = pf, binwidth = 1, 
   color = I('black'), fill = I('#F79420')) +
  scale_x_continuous(limits = c(0, 5), breaks = seq(0, 5, 1))

```

***

### Labeling Plots
Notes:

```{r Labeling Plots}
qplot(x = tenure/365, data = pf, binwidth = 0.25, 
      xlab = 'Number of years using Facebook',
      ylab = 'Number of users in the sample',
   color = I('black'), fill = I('#F79420')) +
  scale_x_continuous(limits = c(0, 5), breaks = seq(0, 5, 1))

ggplot(aes(x = tenure / 365), data = pf) + 
  geom_histogram(color = 'black', fill = '#F79420') + 
  scale_x_continuous(breaks = seq(1, 7, 1), limits = c(0, 7)) + 
  xlab('Number of years using Facebook') + 
  ylab('Number of users in sample')
```

***

### User Ages
Notes:

```{r User Ages}
ggplot(aes(x = age), data = pf) + 
  geom_histogram(color = 'black', fill = '#5760AB', binwidth = 1) + 
  #scale_x_discrete(breaks = seq(0, 121, 10), limits = c(0, 120)) + 
  scale_x_discrete(breaks = seq(0, 121, 10)) +
  xlab('Age') + 
  ylab('Number of users in sample')

```

#### What do you notice?
Response:

***

### The Spread of Memes
Notes:

***

### Lada's Money Bag Meme
Notes:

***

### Transforming Data
Notes:

```{r Transformations}
summary(log10(pf$friend_count + 1))
library(gridExtra)

orig <- ggplot(aes(x = friend_count), data = pf) + 
  geom_histogram(color = 'black', fill = '#5760AB') + 
  #scale_x_discrete(breaks = seq(0, 121, 10)) +
  #scale_x_discrete(limits = c(0, 2000))
  xlab('Friend Count') +
  ylab('Number of users in sample')
  
log_scale <- ggplot(aes(x = log10(friend_count + 1)), data = pf) + 
  #coord_trans(x = "log10") +
  geom_histogram(color = 'black', fill = '#5760AB', binwidth = 0.1) + 
  xlab('Friend Count') + 
  ylab('Number of users in sample, log10 scale') 

log_scale_count <- orig + scale_x_log10()
sqrt_scale_count <- orig + scale_x_sqrt()
grid.arrange(orig, log_scale, log_scale_count, sqrt_scale_count, ncol=2)
```

***

### Add a Scaling Layer
Notes:

```{r Add a Scaling Layer}

```

***


### Frequency Polygons

```{r Frequency Polygons}
qplot(x = friend_count, y = ..count../sum(..count..), 
      data = subset(pf, !is.na(gender)), binwidth = 25, 
      geom = 'freqpoly', color = gender) +
  scale_x_continuous(limits = c(0, 1000), breaks = seq(0, 1000, 50)) 

qplot(x = friend_count, y = ..density.., 
      data = subset(pf, !is.na(gender)), binwidth = 25, 
      geom = 'freqpoly', color = gender) + 
  scale_x_continuous(limits = c(0, 1000), breaks = seq(0, 1000, 50)) 


ggplot(aes(x = friend_count, y = ..count../sum(..count..)), data = subset(pf, !is.na(gender))) +
  geom_freqpoly(aes(color = gender), binwidth=10) + 
  scale_x_continuous(limits = c(0, 1000), breaks = seq(0, 1000, 50)) + 
  xlab('Friend Count') + 
  ylab('Percentage of users with that friend count')

# Note that sum(..count..) will sum across color, so the percentages displayed are percentages of total users. To plot percentages within each group, you can try y = ..density...

ggplot(aes(x = friend_count, y = ..density..), data = subset(pf, !is.na(gender))) +
  geom_freqpoly(aes(color = gender), binwidth=10) + 
  scale_x_continuous(limits = c(0, 1000), breaks = seq(0, 1000, 50)) + 
  xlab('Friend Count') + 
  ylab('Percentage of users with that friend count')
```

***

### Likes on the Web
Notes:

```{r Likes on the Web}
ggplot(aes(x = www_likes + 1, y = ..density..), data = subset(pf, !is.na(gender))) +
  geom_freqpoly(aes(color = gender)) + 
  #geom_histogram() +
  #scale_x_continuous(limits = c(0, 1000), breaks = seq(0, 1000, 50)) + 
  scale_x_log10() + 
  xlab('number of likes') + 
  ylab('Percentage of users with that "likes" count')

qplot(x = www_likes + 1, y = ..density.., 
      data = subset(pf, !is.na(gender)),
      geom = 'freqpoly', color = gender) +
#  scale_x_continuous(limits = c(0, 1000), breaks = seq(0, 1000, 50)) 
  scale_x_log10() 


```


***

### Box Plots
Notes:

```{r Box Plots}

```

#### Adjust the code to focus on users who have friend counts between 0 and 1000.

```{r}

```

***

### Box Plots, Quartiles, and Friendships
Notes:

```{r Box Plots, Quartiles, and Friendships}

```

#### On average, who initiated more friendships in our sample: men or women?
Response:
#### Write about some ways that you can verify your answer.
Response:
```{r Friend Requests by Gender}

```

Response:

***

### Getting Logical
Notes:

```{r Getting Logical}

```

Response:

***

### Analyzing One Variable
Reflection:

***

Click **KnitHTML** to see all of your hard work and to have an html
page of this lesson, your answers, and your notes!